A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-16 tex/body.tex
some edits

\begin{conf}
Detailed proofs of soundness and completeness are included in the technical report
associated with this paper \cite{our-report}.
\end{conf}%
The procedure is implemented in the SMT solver CVC4 as a combination
of rewriting and a theory solver
(Section~\ref{sec:the-theory-solver}).
\rem{CHECK WITH IMPLEMENTATION, but round to nearest 50 or 1000.}
It consists of about 2000 lines of \cpp{} code, among which 1600 are
shared between datatypes and codatatypes. The code is integrated in the
development version of the solver and is expected to be part of the CVC4~1.5 release.
%
An evaluation on %hand-crafted examples and on
problems generated from Isabelle theories using the Sledgehammer tool
demonstrates the usefulness of the approach (Section~\ref{sec:experimental-results}).

@@

To simplify the presentation, we make the following assumptions about
the signature $\Sig$.
First, all codatatypes in $\Codata$ are corecursive. This is not restrictive
because noncorecursive codatatypes can be seen instead as nonrecursive
datatypes.
Second, all types $\tau \in \Nondata$ have infinite cardinality.
This is not overly restrictive, because each interpreted type $\tau \in \Nondata$
having finite cardinality~$n$
can be replaced by an enumeration datatype with $n$~nullary constructors.
Moreover, since the constraints in $\Ec$ are ground, they cannot entail an upper bound
on the cardinality of any uninterpreted type $\tau \in \Nondata$; thus,
we may without loss of generality %or completeness
treat all types in $\Nondata$ as infinite.

@@

The mapping $\Val$ is computed as follows.
%
With each equivalence class $\ec{u}$, we associate a fresh variable $\Varec{u}$
of the same type as $u$
%not occurring in $\tEc$,
and set $\Val\ec{u} = {\Varec{u}}$.
Thus, there are initially no constraints on the values for any equivalence class $\ec{u}$.
The mapping $\Val$ is refined by applying the following unfolding rule exhaustively:\strut
\[
\hbox{\(
\inferrule{
  {\Varec{u}} \in \FV( \Val )
  \betweenantes
  \const{C}( t_1, \ldots, t_n ) \in \ec{u}
  \betweenantes
  \const{C} \in \Ctr
}{
  \Val := \Val [{\Varec{u}} \mapsto \MU {\Varec{u}}.\; \const{C}( {\Varec{t_1}}, \ldots, {\Varec{t_n}} )]
}
\)
%\rn{Unfold}
}
\]
$\FV( \Val )$ denotes the union of the free variables occurring in $\Val$'s range,
and $\Val[x \mapsto t]$ denotes the substitution of $t$ for $x$ in $\Val$'s range.
It is easy to see that the height of terms produced as a result of the unfolding
is bounded by the number of equivalence classes of $\Ec$,
and thus the construction of $\Val$ will terminate.

@@

Given the mapping $\Val$, the \rn{Acyclic} and \rn{Unique} rules operate as follows.
For acyclicity, if $\ec{t\typ{\delta}}$ is a datatype equivalence class
whose values $\Val \ec{t} = \MU x.\; u$ are cyclic
(as indicated by the precondition $x \in \FV( u )$),
then $\Ec$ is unsatisfiable.
For uniqueness, if $\ec{t}$, $\ec{u}$ are two codatatype equivalence classes
whose values $\Val \ec{t}$, $\Val \ec{u}$ are $\alpha$-equivalent,
%for a capture-avoiding renaming of $\mu$-bound variables,
then $t \teq u$. This rule may seem too restrictive, since
$\MU x.\; \const{S}(x)$ and $\MU y.\; \const{S}(\const{S}(y))$ specify the same
value despite being $\alpha$-disequivalent, but the rule will
make some progress by discovering that the subterm $\const{S}(y)$ of $\MU y.\;
\const{S}(\const{S}(y))$ must be equal to the entire term.

@@

\begin{examplex}
Let $\Ec = \{ x \teq \const{S}(x),\; y \teq \const{S}(\const{S}(y)) \}$.
After phase~1, the equivalence classes %of $\Ec$
are
$\{ x,\, \const{S}(x) \}$,
$\{ y,\, \const{S}(\const{S}(y)) \}$,
and $\{ \const{S}(y) \}$.
Constructing $\Val$ yields
\begin{align*}
\Val \ec{x} & \,=\, \MU {\Varec{x}}.\; \const{S}(\Varec{x}) &
\enskip %% TYPESETTING
\Val \ec{y} & \,=\,
\MU {\Varec{y}}.\; \const{S}(\smash{\MU {\VAREC{\const{S}(y)}}}.\; \const{S}( {\Varec{y}} ) )
\enskip %% TYPESETTING
 &
\Val \ec{\const{S}(y)} & =
\MU {\VAREC{\const{S}(y)}}.\; \const{S}(\MU {\Varec{y}}.\; \const{S}( {\VAREC{\const{S}(y)}} ) )
\end{align*}
Since $\Val \ec{y} \aequiv \Val \ec{\const{S}(y)}$,
%for the renaming
%$\{ {\Varec{y}} \mapsto {\VAREC{\const{S}(y)}}, {\VAREC{\const{S}(y)}} \mapsto {\Varec{y}} \}$,
the \rn{Unique} rule applies to derive $y \teq \const{S}(y)$.
%Intuitively, this equality holds, since the values of $y$ and $\const{S}(y)$
%are equivalent.
%
Next, phase~1 is activated again to produce the equivalence classes
$\{ x,\, \const{S}(x) \}$ and
$\{ y,\, \const{S}(y),\, \const{S}(\const{S}(y)) \}$.
The mapping $\Val$ is updated accordingly:
\begin{conf}%
$\Val \ec{y} \,=\, \MU {\Varec{y}}.\; \const{S}({\Varec{y}})$.
\end{conf}%
\begin{rep}
\begin{align*}
\Val \ec{x} & \,=\, \MU {\Varec{x}}.\; \const{S}(\Varec{x})
&
\Val \ec{y} & \,=\, \MU {\Varec{y}}.\; \const{S}({\Varec{y}})
\end{align*}
\end{rep}
Finally, since $\Val \ec{x} \aequiv \Val \ec{y}$,
%for the renaming
%$\{ {\Varec{y}} \mapsto {\VAREC{\const{S}(y)}}, {\VAREC{\const{S}(y)}} \mapsto {\Varec{y}} \}$,
the \rn{Unique} rule derives $x \teq y$.
\xend
\end{examplex}

@@

\paragraph{Phase 3: Branching.}
If a selector is applied to a term $t$, or if $t$'s type is a finite datatype,
the equivalence class of $t$ must contain a
$\delta$ constructor term.
This is enforced in the third phase by the \rn{Split} rule.
A companion rule, \rn{Single}, focuses on the degenerate case where two
terms from $\tEc$ have a corecursive singleton type
(Section~\ref{sec:the-theory-of-co-datatypes}) and must therefore be equal. Both
the finiteness assumption on the datatype in \rn{Split} and the singleton
constraint on the codatatype in \rn{Single} can be computed statically
based on a recursive computation of the cardinalities of the
constructors' argument types (and assuming all types in $\Nondata$ are
infinite).

>>>>

\begin{conf}
Detailed proofs of soundness and completeness are included in the technical report
associated with this paper \cite{our-report}.
\end{conf}%
The procedure is implemented in the SMT solver CVC4 as a combination
of rewriting and a theory solver
(Section~\ref{sec:the-theory-solver}).
\rem{CHECK WITH IMPLEMENTATION, but round to nearest 50 or 100 LOC.}
It consists of about 2000 lines of \cpp{} code, among which 1600 are
shared between datatypes and codatatypes. The code is integrated in the
development version of the solver and is expected to be part of the CVC4~1.5 release.
%
An evaluation on %hand-crafted examples and on
problems generated from Isabelle theories using the Sledgehammer tool
demonstrates the usefulness of the approach (Section~\ref{sec:experimental-results}).

@@

To simplify the presentation, we make the following assumptions about
the signature $\Sig$.
First, all codatatypes in $\Codata$ are corecursive. This is not restrictive
because noncorecursive codatatypes can be seen instead as nonrecursive
datatypes.
Second, all types $\tau \in \Nondata$ have infinite cardinality.
This is not overly restrictive, because each interpreted type $\tau \in \Nondata$
having finite cardinality~$n$
can be replaced by an enumeration datatype with $n$~nullary constructors.
Finally, since the constraints in $\Ec$ are ground, they cannot entail an upper
bound on the cardinality of any uninterpreted type $\tau \in \Nondata$; thus,
we may without loss of generality %or completeness
treat all types in $\Nondata$ as infinite.

@@

The mapping $\Val$ is computed as follows.
%
With each equivalence class $\ec{u}$, we associate a fresh variable $\Varec{u}$
of the same type as $u$
%not occurring in $\tEc$,
and let $\Val\ec{u} = {\Varec{u}}$. Therefore, there are initially no
constraints on the values for any equivalence class $\ec{u}$. The mapping $\Val$
is refined by applying the following unfolding rule exhaustively:\strut
\[
\hbox{\(
\inferrule{
  {\Varec{u}} \in \FV( \Val )
  \betweenantes
  \const{C}( t_1, \ldots, t_n ) \in \ec{u}
  \betweenantes
  \const{C} \in \Ctr
}{
  \Val := \Val [{\Varec{u}} \mapsto \MU {\Varec{u}}.\; \const{C}( {\Varec{t_1}}, \ldots, {\Varec{t_n}} )]
}
\)
%\rn{Unfold}
}
\]
$\FV( \Val )$ denotes the union of the free variables occurring in $\Val$'s range,
and $\Val[x \mapsto t]$ denotes the substitution of $t$ for $x$ in $\Val$'s range.
It is easy to see that the height of terms produced as a result of the unfolding
is bounded by the number of equivalence classes of $\Ec$,
and thus the construction of $\Val$ will terminate.

@@

Given the mapping $\Val$, the \rn{Acyclic} and \rn{Unique} rules operate as follows.
For acyclicity, if $\ec{t}$ is a datatype equivalence class
whose values $\Val \ec{t} = \MU x.\; u$ are cyclic
(as required by the precondition $x \in \FV( u )$),
then $\Ec$ is unsatisfiable.
For uniqueness, if $\ec{t}$, $\ec{u}$ are two codatatype equivalence classes
whose values $\Val \ec{t}$, $\Val \ec{u}$ are $\alpha$-equivalent,
%for a capture-avoiding renaming of $\mu$-bound variables,
then $t \teq u$. This may seem too restrictive, since
$\MU x.\; \const{S}(x)$ and $\MU y.\; \const{S}(\const{S}(y))$ specify the same
value despite being $\alpha$-disequivalent, but the rule will
make some progress by discovering that the subterm $\const{S}(y)$ of $\MU y.\;
\const{S}(\const{S}(y))$ must be equal to the entire term.

@@

\begin{examplex}
Let $\Ec = \{ x \teq \const{S}(x),\; y \teq \const{S}(\const{S}(y)) \}$.
After phase~1, the equivalence classes %of $\Ec$
are
$\{ x,\, \const{S}(x) \}$,
$\{ y,\, \const{S}(\const{S}(y)) \}$,
and $\{ \const{S}(y) \}$.
Constructing $\Val$ yields
\begin{align*}
\Val \ec{x} & \,=\, \MU {\Varec{x}}.\; \const{S}(\Varec{x}) &
\enskip %% TYPESETTING
\Val \ec{y} & \,=\,
\MU {\Varec{y}}.\; \const{S}(\smash{\MU {\VAREC{\const{S}(y)}}}.\; \const{S}( {\Varec{y}} ) )
\enskip %% TYPESETTING
 &
\Val \ec{\const{S}(y)} & =
\MU {\VAREC{\const{S}(y)}}.\; \const{S}(\MU {\Varec{y}}.\; \const{S}( {\VAREC{\const{S}(y)}} ) )
\end{align*}
Since $\Val \ec{y} \aequiv \Val \ec{\const{S}(y)}$,
%for the renaming
%$\{ {\Varec{y}} \mapsto {\VAREC{\const{S}(y)}}, {\VAREC{\const{S}(y)}} \mapsto {\Varec{y}} \}$,
the \rn{Unique} rule applies to derive $y \teq \const{S}(y)$.
%Intuitively, this equality holds, since the values of $y$ and $\const{S}(y)$
%are equivalent.
%
At this point, phase~1 is activated again to refine the equivalence classes,
yielding
$\{ x,\, \const{S}(x) \}$ and
$\{ y,\, \const{S}(y),\allowbreak\, \const{S}(\const{S}(y)) \}$.
The mapping $\Val$ is updated accordingly:
\begin{conf}%
$\Val \ec{y} \,=\, \MU {\Varec{y}}.\; \const{S}({\Varec{y}})$.
\end{conf}%
\begin{rep}
\begin{align*}
\Val \ec{x} & \,=\, \MU {\Varec{x}}.\; \const{S}(\Varec{x})
&
\Val \ec{y} & \,=\, \MU {\Varec{y}}.\; \const{S}({\Varec{y}})
\end{align*}
\end{rep}
Since $\Val \ec{x} \aequiv \Val \ec{y}$,
%for the renaming
%$\{ {\Varec{y}} \mapsto {\VAREC{\const{S}(y)}}, {\VAREC{\const{S}(y)}} \mapsto {\Varec{y}} \}$,
the \rn{Unique} rule can finally derive $x \teq y$.
\xend
\end{examplex}

@@

\paragraph{Phase 3: Branching.}
If a selector is applied to a term $t$, or if $t$'s type is a finite datatype,
the equivalence class of $t$ must contain a
$\delta$ constructor term.
This is enforced in the third phase by the \rn{Split} rule.
A companion rule, \rn{Single}, focuses on the degenerate case where two
terms from $\tEc$ have a corecursive singleton type
(Section~\ref{sec:the-theory-of-co-datatypes}) and must therefore be equal. Both
the finiteness assumption on the datatype in \rn{Split} and the singleton
constraint on the codatatype in \rn{Single} can be computed statically
based on a recursive computation of the cardinalities of the
constructors' argument types (assuming all types in $\Nondata$ are
infinite).
